---
title: "Data Modeling"
author: "Owen Simon"
format: 
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Contents"
    embed-resources: true
execute:
  include: true
  eval: true    
  warning: false
  message: false
---

# 1. Setup

```{r}
# Load libraries
library(tidyverse)
library(pROC)
library(glmnet)
library(randomForest)
library(gt)
library(doParallel)
library(caret)
library(tibble)

options(na.action = na.pass)

data_dir <- "data-files"

# Read datasets
train <- read_csv(file.path(data_dir, "train_final.csv")) |>
  mutate(across(where(is.character), as.factor))

test <- read_csv(file.path(data_dir, "test_final.csv")) |>
  mutate(across(where(is.character), as.factor))
```

# 2. Majority Class Model

## 2.1 Overview

```{r}
TARGET_dist <- train |> 
  count(TARGET) |>
  mutate(proportion = n / sum(n))

gt(TARGET_dist)
```

The target variable is highly imbalanced, with the majority class (TARGET = 0) representing approximately 91.9% of the training data. This imbalance will be an important consideration when evaluating model performance.

## 2.2 Baseline Model Performance

```{r}
# Baseline predictions (majority class)
baseline_predictions <- rep.int(0, nrow(train))
baseline_accuracy <- mean(baseline_predictions == train$TARGET)

# Baseline probabilities (probability of TARGET = 1)
majority_class_prob <- TARGET_dist$proportion[TARGET_dist$TARGET == 0]
baseline_probs <- rep(majority_class_prob, nrow(train))

baseline_auc <- roc(train$TARGET, baseline_probs)$auc
baseline_results <- tibble(
  Metric = c("Baseline Accuracy", "Baseline AUC"),
  Value = c(baseline_accuracy, as.numeric(baseline_auc))
)

gt(baseline_results)
```

A baseline model that always predicts the majority class would achieve an accuracy of around 91.9%. However, this model would have an AUC of 0.5, as it does not differentiate between the classes.

# 3. Compare Candidate Models

## 3.1 Data Preparation
```{r}
raw_app_train <- read_csv(file.path(data_dir, "application_train.csv")) |>
  mutate(across(where(is.character), as.factor))

# Keep only application-based predictors
application_cols <- colnames(raw_app_train)
application_cols_in_train <- intersect(application_cols, colnames(train))

train_app <- train |>
  select(all_of(application_cols_in_train), TARGET, -SK_ID_CURR) |>
  mutate(TARGET = factor(TARGET, levels = c(0,1)))
```

The initial model comparison will focus on the original application data. Supplementary data sources (bureau, previous applications, etc.) will be explored in future iterations.

## 3.2 Cross-Validation Control

```{r}
set.seed(42)
n_folds <- 5
fold_ids <- sample(rep(1:n_folds, length.out = nrow(train_app)))

# Storage for fold predictions
preds_lr <- numeric(nrow(train_app))
preds_lasso <- numeric(nrow(train_app))
preds_rf <- numeric(nrow(train_app))

# Storage for number of predictors
n_pred_lr <- n_pred_lasso <- n_pred_rf <- NA
```

Five fold cross validation is used to estimate out of sample model performance. For each fold, models are trained on four fifths of the training data and evaluated on the remaining one fifth. This process ensures that performance metrics reflect generalization ability rather than training set fit.

## 3.3 Parallel Processing

```{r}
n_cores <- 2
cl <- makePSOCKcluster(n_cores)
registerDoParallel(cl)
```

## 3.4 Modeling Folds

```{r}
for (k in 1:n_folds) {
  cat("Processing fold", k, "\n")
  
  train_fold <- train_app[fold_ids != k, ]
  test_fold  <- train_app[fold_ids == k, ]
  
  # Logistic Regression
  fit_lr <- glm(TARGET ~ ., data = train_fold, family = binomial)
  preds_lr[fold_ids == k] <- predict(fit_lr, test_fold, type = "response")
  if (is.na(n_pred_lr)) n_pred_lr <- length(coef(fit_lr)) - 1
  
  # LASSO Logistic Regression
  x_train <- model.matrix(TARGET ~ . - 1, data = train_fold)
  y_train <- as.numeric(train_fold$TARGET) - 1
  x_test <- model.matrix(TARGET ~ . - 1, data = test_fold)
  
  lasso <- cv.glmnet(x_train, y_train, family = "binomial", alpha = 1, parallel = TRUE)
  pred_lasso_fold <- predict(lasso, x_test, s = "lambda.min", type = "response")
  preds_lasso[fold_ids == k] <- pred_lasso_fold
  if (is.na(n_pred_lasso)) {
    coefs <- coef(lasso, s = "lambda.min")
    n_pred_lasso <- sum(coefs[-1, ] != 0)
  }
  
  # Random Forest
  set.seed(42 + k)
  rf <- randomForest(TARGET ~ ., data = train_fold, ntree = 200)
  pred_rf_fold <- predict(rf, test_fold, type = "prob")[, "1"]
  preds_rf[fold_ids == k] <- pred_rf_fold
  if (is.na(n_pred_rf)) n_pred_rf <- ncol(train_fold) - 1
}
```

## 3.5 Stop Parallel Cluster

```{r}
stopCluster(cl)
registerDoSEQ()
```

## 3.6 Model Comparison

### 3.6.1 Compute CV-AUC

```{r}
auc_lr <- auc(roc(train_app$TARGET, preds_lr))
auc_lasso <- auc(roc(train_app$TARGET, preds_lasso))
auc_rf <- auc(roc(train_app$TARGET, preds_rf))
```

The reported AUC values are cross validated estimates computed from out of fold predictions. Because each observation is evaluated using a model that was not trained on it, these values provide an unbiased estimate of expected performance on unseen data.

### 3.6.2 Candidate Model Summary Table

```{r}
comparison_table <- tibble(
  Model = c("Logistic", "LASSO", "Random Forest"),
  CV_AUC = c(auc_lr, auc_lasso, auc_rf),
  Num_Predictors = c(n_pred_lr, n_pred_lasso, n_pred_rf)
)

gt(comparison_table)
```

The LASSO model will be used for further development as it achieves the highest cross-validated AUC among candidate models while retaining fewer predictors than standard Logistic Regression.

# 4. Model Optimization

## 4.1 Add Supplementary Features

```{r}
# full training data (application + supplementary features)
train_model <- train |> select(-SK_ID_CURR)

# recode TARGET levels
train_model$TARGET <- factor(train_model$TARGET, levels = c(0,1), labels = c("No","Yes"))
```

## 4.2 Fit LASSO with SMOTE

```{r}
# Define a grid for lambda (regularization) and alpha (mix between lasso/ridge)
grid <- expand.grid(
  alpha = seq(0, 1, by = 0.2),
  lambda = 10^seq(-4, 0, length = 20)
)

ctrl <- trainControl(
  method = "cv",
  number = 5,
  sampling = "smote",            # handle class imbalance
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  verboseIter = TRUE
)

set.seed(42)
lasso_caret <- train(
  TARGET ~ .,
  data = train_model,
  method = "glmnet",
  family = "binomial",
  metric = "ROC",
  trControl = ctrl,
  tuneGrid = grid
)
```

Hyperparameter tuning explores combinations of alpha and lambda using five fold cross validation on the training data. The value of alpha controls the balance between ridge and LASSO penalties, while lambda controls the overall strength of regularization. The optimal combination is selected based on cross validated ROC AUC. After tuning is complete, the final model is automatically refit on the full training dataset using the selected hyperparameters.

SMOTE is applied within each cross validation fold to address class imbalance. Synthetic minority class observations are generated only within the training portion of each fold, preventing information leakage into validation data and improving the model’s ability to distinguish between classes.

## 4.3 Final Model Features

The final tuned LASSO model is refit on the full training dataset using the optimal hyperparameters selected during cross validation. The resulting coefficient estimates determine which predictors are retained in the final model.

### 4.3.1 Number of Predictors and Selected Features
```{r}
# Extract coefficients at best lambda
coef_lasso <- coef(lasso_caret$finalModel, s = lasso_caret$bestTune$lambda)

# Number of predictors (non-zero coefficients)
num_predictors <- sum(coef_lasso != 0) - 1  # subtract intercept
num_predictors

# Selected predictor names
selected_vars <- rownames(coef_lasso)[coef_lasso[,1] != 0]
selected_vars <- selected_vars[selected_vars != "(Intercept)"]

# Display selected features in a table
selected_vars_table <- tibble(
  Selected_Feature = selected_vars
)
```

### 4.3.2 Selected Features Table

```{r}
gt(selected_vars_table) |>
  tab_header(
    title = "Features Selected by Final LASSO Model"
  )
```

## 4.4 Cross-Validation AUC

```{r}
cv_auc <- max(lasso_caret$results$ROC)
cv_auc
```

The maximum cross-validated ROC AUC across hyperparameter combinations is 0.7688, indicating the expected discrimination performance of the tuned LASSO model on unseen data.

## 4.5 Predict on Test Set (Using Best Hyperparameters)

```{r}
# Prepare test data (exclude ID column)
test_model <- test |> select(-SK_ID_CURR)

# Predicted probabilities for TARGET = "Yes"
pred_test_probs <- predict(lasso_caret, newdata = test_model, type = "prob")[, "Yes"]
```

Predictions on the test set are generated using the final tuned model obtained from cross validation. The test data are not used during model training or hyperparameter selection and serve only as new unseen observations for which predicted probabilities are required.

# 5. Create Submission File
```{r}
submission_full <- tibble(
  SK_ID_CURR = test$SK_ID_CURR,
  TARGET = pred_test_probs
)

# Write submission CSV
write_csv(submission_full, "submission_lasso_full.csv")
```

The resulting submission file received a Kaggle area under the ROC curve of 0.76187 private and 0.77041 public. The private leaderboard is calculated with approximately 80% of the test data.