---
title: "Credit Default Prediction Model Card"
author: "Owen Simon"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
# load libraries
library(tidyverse)
library(pROC)
library(caret)
```

# Executive Summary

This model predicts the probability that a loan applicant will default on a consumer loan. The model was trained using the Home Credit dataset and optimized using LASSO logistic regression.

The model’s purpose is to help lenders evaluate borrower risk and improve credit allocation decisions.

Using cross-validated predictions, the model achieves strong discrimination between borrowers who repay and those who default. Based on realistic lending assumptions about loan profitability and losses from default, an optimal probability threshold was selected to maximize expected financial return.

At this threshold:

Expected approval rate: XX%

Expected portfolio default rate: XX%

Estimated improvement in portfolio profit relative to approving all borrowers: XX%

The model should be used as a decision support tool rather than the sole basis for lending decisions. Human review and regulatory compliance processes remain necessary.

Key caveats include potential bias in historical credit data, limited visibility into informal income sources, and potential changes in borrower behavior over time.

# 1 - Model Details

## 1.1 - Model Type
- Algorithm: LASSO Logistic Regression (`glmnet`)
- Implementation: `glmnet` via `caret` package
- Training method: 5-fold cross-validation
- Objective: Predict probability of loan default (`TARGET` = 1)

```{r}
lasso_final <- readRDS("lasso_final_model.rds")
feature_names <- readRDS("feature_names.rds")
coef_df <- readRDS("lasso_feature_importance.rds")
```

## 1.2 - Key Hyperparameters

```{r}
lasso_final$bestTune
```

## 1.3 - Training Data Summary

The model was trainined on the **Home Credit Default Risk** dataset, which contains 198 unique borrower demographic, financial, and credit history variables.

- Target variable: `TARGET` (1 = default, 0 = no default)
- Dataset Characteristics:
  - Highly imbalanced
  - Loan default rate of 8.1%

The dataset includes engineered features dereived from supplementary credit bureau and previous loan application tables.

# 2 - Intended Use

## 2.1 - Intended Users

## 2.2 - Decisions Supported

## 2.3 - Out of Scope

# 3 - Performance Metrics

```{r}
val_model <- readRDS("validation_data.rds")
val_probs <- readRDS("validation_probs.rds")

roc_obj <- roc(val_model$TARGET, val_probs)
val_auc <- auc(roc_obj)

val_auc
```

## 3.1 - Cross-Validated Performance

```{r}
cv_auc <- max(lasso_final$results$ROC)
cv_auc
```

## 3.2 - Precision and Recall

```{r}
threshold <- 0.5
pred_class <- ifelse(val_probs > threshold,"Yes","No")
pred_class <- factor(pred_class,levels=c("No","Yes"))

confusionMatrix(pred_class,val_model$TARGET)
```