---
title: "Model Card: Home Credit Default Risk Model"
author: "Owen Simon"
format:
  html:
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
lasso_final <- readRDS("lasso_final_model.rds")
val_model <- readRDS("validation_data.rds")
val_probs <- readRDS("validation_probs.rds")
feature_names <- readRDS("feature_names.rds")
coef_df <- readRDS("lasso_feature_importance.rds")
```

# Executive Summary

Home Credit aims to responsibly expand lending to applicants without traditional credit histories. This predictive model estimates the probability that each applicant will successfully repay a loan, leveraging alternative data sources, including application information and telecommunications data.

**Business Recommendation:** Approve loans for applicants with repayment probabilities above the chosen threshold; deny or adjust loans for those below it.  

**Expected Financial Impact:** At the selected threshold, estimated profit per successfully repaid loan is \$600, while expected losses from defaults average 45% of the loan amount, with recovery rates around 25%. The model balances expanding access to credit while minimizing risk.  

**Key Caveats:** Accuracy depends on alternative data availability and quality. Economic changes, policy shifts, or missing borrower information may impact predictions. Monitor approval disparities across demographic groups.

---

# Model Details

- **Type:** LASSO Logistic Regression (`glmnet`)
- **Version:** v1.0
- **Training Data:**
  - Source: Home Credit application and alternative data
  - Rows: `r nrow(train_model)`
  - Features: `r ncol(train_model) - 1` (excluding target)
  - Key Features: EXT_SOURCE_1, AMT_CREDIT, DAYS_EMPLOYED, NAME_CONTRACT_TYPE, CODE_GENDER, NAME_EDUCATION_TYPE
- **Hyperparameters:**
  - L1/L2 regularization determined via `tuneGrid = best_no`
  - Family: Binomial (logistic regression)
  - Cross-validation: 5-fold (`trainControl(method = "cv", number = 5, classProbs = TRUE, summaryFunction = twoClassSummary)`)

**Note:** Model trained to maximize ROC AUC with stratified 5-fold CV.

---

# Intended Use

- **Users:** Credit risk analysts, lending officers
- **Decisions Informed:** Loan approval, loan amount, loan term
- **Not Designed For:** Predicting other financial behaviors or for populations outside the model’s training distribution

---

# Performance Metrics

- **Cross-Validated AUC:** 0.7562
- **Kaggle Test Set AUC:**  
  - Public: 0.75230
  - Private: 0.74213

- **Precision / Recall at chosen threshold:** *(Insert values based on threshold)*  
- **Additional Metrics:** Confusion matrix and F1 score can be derived from selected threshold

```{r}
# Example placeholder for ROC curve
# plot(roc(train_model$TARGET, predict(lasso_final, train_model, type="prob")[,2]))