---
title: "Credit Default Prediction Model Card"
author: "Owen Simon"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
# load libraries
library(tidyverse)
library(pROC)
library(caret)
library(gt)
library(dplyr)
```

```{r}
cv_preds <- readRDS("rds-files/cv_preds_lasso_full.rds")
val_model <- readRDS("rds-files/validation_data.rds")
val_probs <- readRDS("rds-files/validation_probs.rds")
```

# Executive Summary

This model predicts the probability that a borrower will default on a consumer loan. It was trained using loan application data, combined with supplemental datasets detailing historical credit behavior.

The model supports lenders in evaluating borrower risk and improving credit allocation decisions. Using cross-validated predictions, the model discriminates well between borrowers who are likely to repay and those at high risk of default.

Based on realistic assumptions about loan profitability and losses from default, an optimal probability threshold was selected to maximize expected financial return.

**At the profit-maximizing threshold:**
- **Approval Rate:**
- **Portfolio default rate:**
- **Estimated portfolio profit:**

The model is intended as a **decision support tool**, not a replacement for human review or regulatory compliance.

**Key caveats:** Potential bias in historical data, incomplete visibility into informal income, and changes in borrower behavior over time.

# 1) Model Details

This section describes the **model type, version, training data, and key hyperparameters**.

## 1.1 - Model Type

- **Algorithm:** LASSO Logistic Regression (`glmnet`)  
- **Implementation:** `glmnet` via the `caret` package  
- **Training method:** 5-fold cross-validation with SMOTE sampling  
- **Objective:** Predict probability of loan default (`TARGET` = 1)  

```{r}
# Load final model and feature importance
lasso_final <- readRDS("rds-files/lasso_final_model.rds")
feature_names <- readRDS("rds-files/feature_names.rds")
coef_df <- readRDS("rds-files/lasso_feature_importance.rds")
```

### 1.2 - Model Version

| Attribute | Value |
|----------|------|
| Model name | Credit Default Prediction Model |
| Algorithm | LASSO Logistic Regression |
| Implementation | `glmnet` via `caret` |
| Training procedure | 5-fold cross-validation |
| Target variable | `TARGET` (1 = default, 0 = repaid) |
| Output | Probability of loan default |

### 1.3 - Training Data Summary

The model was trained on **loan application data** with supplemental credit history datasets. Features were aggregated at the borrower level to produce a comprehensive predictor set.

| Dataset | Description |
|---------|-------------|
| `application_train` | Loan applications with demographic, financial, and target information |
| `application_test` | Loan applications without the target variable, used for evaluation |
| `bureau` | Historical credit bureau records |
| `bureau_balance` | Monthly loan balance status from bureau reports |
| `previous_application` | Records of previous loan applications |
| `credit_card_balance` | Credit card balances and repayment behavior |
| `POS_CASH_balance` | Point-of-sale and cash loan repayment history |
| `installments_payments` | Historical installment payment records |

**Target variable:**

- `TARGET = 1` → borrower defaulted  
- `TARGET = 0` → borrower repaid  

The dataset is **highly imbalanced**, with defaults representing a small proportion of all observations.

### 1.4 - Key Hyperparameters

```{r}
# Display the best hyperparameters for the final model
lasso_final$bestTune
```

These hyperparameters were selected using cross-validation on a subset of the training data and SMOTE sampling.

## 2) - Intended Use

### 2.1 - Intended Users

This model is designed for use by:

- Credit risk analysts evaluating borrower applications  
- Consumer lending decision systems and automated underwriting pipelines  
- Risk management teams monitoring portfolio performance  

### 2.2 - Decisions Supported

The model supports decisions such as:

- Approving or denying loan applications based on predicted default risk  
- Identifying borrowers who may require additional review or manual underwriting  
- Prioritizing applications for follow-up or risk-based pricing adjustments  

### 2.3 - Out of Scope

This model is **not intended to:**

- Predict repayment amounts or timing beyond default/no-default
- Replace human judgment in underwriting decisions  
- Predict defaults in a different population or market without retraining  
- Be used outside consumer lending contexts

# 3) Performance Metrics

This section evaluates model performance using **Area Under the ROC Curve (AUC)**.  
AUC is a threshold-independent metric that measures how well the model separates borrowers who default from those who repay.

Metrics that depend on a classification cutoff (such as **precision, recall, approval rates, and confusion matrices**) will be evaluated after selecting an optimal decision threshold in Section 4.

## 3.1 Cross-Validated Performance

The final model was trained using **5-fold cross-validation with SMOTE sampling** to address class imbalance.  
The cross-validated AUC provides an estimate of model performance during training.

```{r}
cv_auc <- readRDS("rds-files/cv_auc_lasso_full.rds")
cv_auc
```

## 3.2 - Validation Set Performance

To obtain an unbiased estimate of performance, predictions were generated on a hold-out validation set that was not used during training.

```{r}
val_model <- readRDS("rds-files/validation_data.rds")
val_probs <- readRDS("rds-files/validation_probs.rds")

roc_obj <- roc(val_model$TARGET, val_probs)
auc_val <- auc(roc_obj)

auc_val
```

The validation AUC reflects the model’s ability to distinguish between defaulting and non-defaulting borrowers on unseen data.

## 3.3 - Performance Summary Table

The table below compares the cross-validated AUC from training with the validation AUC from the hold-out dataset.

```{r}
auc_table <- tibble(
  Metric = c("Cross-Validated AUC", "Validation AUC"),
  Value  = c(cv_auc, auc_val)
)

auc_table |>
  mutate(Value = round(Value, 4)) |>
  gt() |>
  cols_label(
    Metric = "Metric",
    Value = "Value"
  ) |>
  fmt_number(
    columns = Value,
    decimals = 4
  )
```

The cross-validated AUC and validation AUC are similar,
indicating that the model generalizes reasonably well to unseen data and does
not show strong evidence of overfitting.

## 3.4 - Kaggle Benchmark

For additional context, model performance can be compared to scores from the
*Home Credit Default Risk* Kaggle competition.

| Benchmark | AUC |
|---|---|
| Cross-Validated AUC | 0.7693 |
| Validation AUC | 0.7520 |
| Kaggle Public Score | 0.7418 |
| Kaggle Private Score | 0.7304 |

The Kaggle leaderboard scores are slightly lower than the cross-validated and
validation AUC values. This is expected because the Kaggle test set represents
completely unseen data, while the validation set is drawn from the same
distribution as the training data. Minor differences also arise because the
private leaderboard is calculated using approximately 80% of the test set.